---
title: "Midterm Study"
author: "Steve Bramhall"
date: "June 27, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tswge)
```

Plot ACF's, log and non-log spectral densities
```{r Prob 1.5}
data(fig1.21a)                       # load data
plotts.sample.wge(fig1.21a)          # view sample realization, autocorrelations, peridogram, parzen window
acf(fig1.21a)                        # acf with CIs
pacf(fig1.21a)                       # partial acf, will review after midterm
parzen.wge(fig1.21a,plot=T)          # trunc =2(sqrt(n))
parzen.wge(fig1.21a,plot=T,dbcalc=F) # no decible scale, log accentuates the peaks
parzen.wge(fig1.21a,plot=T,trunc=31) # trunc M = 21
```
Plot signal plus noise
```{r Prob 1.6}
# x = 3cos(2pi(0.05)t) + 1.5cos(2pi(0.35)t + 2) +at
x = gen.sigplusnoise.wge(n=100,coef=c(3,1.5),freq=c(.05,.35),psi=c(0,2)) # signal + noise
plotts.sample.wge(x)                                                     # sample autocorrelations
```
Plot with filtering, butterworth
```{r Prob 2.1}
x = gen.sigplusnoise.wge(n=100,coef=c(3,1.5),freq=c(.05,.35),psi=c(0,2)) # signal + noise
x1 = butterworth.wge(x,order=3,cutoff=.2,type="low")                     # 3rd order low pass w/ cutoff at .2
x2 = butterworth.wge(x,order=3,cutoff=.2,type="high")                    # 3rd order high pass w/ cutoff at .2
x3 = butterworth.wge(x2$x.filt,order=3,cutoff=.2,type="low")             # 3rd order low pass w/ cutoff at .2 on x2

plotts.sample.wge(x)
plotts.sample.wge(x1$x.filt)
plotts.sample.wge(x2$x.filt)
plotts.sample.wge(x3$x.filt)
```
Plot with differencing filter
```{r Prob 2.3 Diff Filter}
x1 = gen.sigplusnoise.wge(n=100,coef=c(3,1.5),freq=c(.05,.35),psi=c(0,2)) # signal + noise
xdif=artrans.wge(x1,phi.tr=1)                                             # apply 1 diff filter
plotts.sample.wge(xdif)                                                   # plot after 1 dif
```
Plot MA filter
```{r Prob 2.3 MA Filter}
x = gen.sigplusnoise.wge(n=100,coef=c(3,1.5),freq=c(.05,.35),psi=c(0,2),sn=8) # signal + noise
ma=filter(x,rep(1/5,5))
plot(ma)
```
Plot MA and AR, get mean and variance, see lecture slide 127 for predicted mean and variance
```{r Prob 3.1 MA and AR Plotts Generation}
# Xt - 25 = at - .95at-1 +.9at-2 -.855at-3
# Xt - 25 = at(1-.95B + .9B^2 - .855B^2)
x=plotts.true.wge(n=200,theta=c(.95,-.9,.855))            # MA(3)
mean(x$data)
var(x$data)
# Xt - .95Xt-1 = at
plotts.true.wge(n=200,phi=.95)                            # AR(1)

# (Xt - 10) - 1.5(Xt-1 - 10) + .9(Xt-2 - 10) = at
# Xt - 10 -1.5Xt-1 + 15 +.9Xt-2 -.9 = at
# 1 - 1.5B + .9B^2 = at - 4.1                             # AR(2)
plotts.true.wge(n=200,phi=c(1.5,-.9))

```
Use factor to check for stationarity
```{r Prob 3.3}
# Are these stationary
plotts.true.wge(phi=c(1.55,-1,.25))
factor.wge(phi=c(1.55,-1,.25))             # Stationary
plotts.true.wge(phi=c(2,-1.7,1.6,-.77))    # Not Stationary
factor.wge(phi=c(2,-1.7,1.6,-.77))
plotts.true.wge(theta=c(2,-1.76,1.6,-.77)) # variance is not constant over time
factor.wge(phi=c(2,-1.76,1.6,-.77))        # Not Stationary
plotts.true.wge(phi=c(1.9,-2.3,2,-1.2,.4))
factor.wge(phi=c(1.9,-2.3,2,-1.2,.4))      # Stationary
```
Plot true and sample plots
```{r Prob 3.4}
# TRUE PLOTS - AUTCORRELATIONS, DENSITY
plotts.true.wge(n=200,phi=c(.9))
plotts.true.wge(n=200,phi=c(-.9))
plotts.true.wge(n=200,phi=c(.5))
plotts.true.wge(n=200,phi=c(-.5))
plotts.true.wge(n=200,phi=c(.1))
plotts.true.wge(n=200,phi=c(-.1))

# REALIZATION PLOTS
x1=gen.arma.wge(n=200,phi=c(.9))
x2=gen.arma.wge(n=200,phi=c(-.9))
x3=gen.arma.wge(n=200,phi=c(.5))
x4=gen.arma.wge(n=200,phi=c(-.5))
x5=gen.arma.wge(n=200,phi=c(.1))
x6=gen.arma.wge(n=200,phi=c(-.1))

# SAMPLE PLOTS - AUTOCORRELATIONS, DENSITY
plotts.sample.wge(x1)
plotts.sample.wge(x2)
plotts.sample.wge(x3)
plotts.sample.wge(x4)
plotts.sample.wge(x5)
plotts.sample.wge(x6)
```
Identify the model by reducing then determine if stationary and invertible
```{r Prob 3.6}
# Xt - Xt-1 + .26Xt-2 + .64Xt-3 - .576Xt-4 = at - 2.4at-1 + 2.1at-2 - .72at-3
factor.wge(phi=c(1,-.26,-.64,.576))  # 1st factor AR part
factor.wge(phi=c(2.4,-2.18,.72))     # then factor MA part and review cancellation opportunities
# The result is (1+.8B) and (1-.8B) cancel each other out so the eqn becomes:
# (1-B+.9B^2)(1-.8B)Xt = (1-1.6B+.9B^2)at  ---> ARM(3,2) ---> p = 3, q = 2
# All Abs Recip are < 1 so the model is invertible

```
Identify the model by reducing then determine if stationary and invertible
```{r Prob 3.7a}
# Xt - .1Xt-1 + .5Xt-2 + .08Xt-3 -.24Xt-4 = at
plotts.true.wge(n=200,phi=c(.1,-.5,-.08))
factor.wge(phi=c(.1,-.5,-.08))  # AR(3), stationary and invertible
```

```{r Prob 3.7b}
# Xt - 1.3Xt-1 + .4Xt-2 = at - 1.9at-1
plotts.true.wge(n=200,phi=c(1.3,-.4),theta=c(1.9)) # stationary
factor.wge(phi=c(1.3,-.4))
factor.wge(phi=c(1.9))                             # not invertible
```

```{r Prob 3.7c}
# Xt - 1.9Xt-1 = at - 1.3at-1 + .4at-2
#plotts.true.wge(n=200,phi=c(1.9),theta=c(1.3,-.4)) # not stationary, won't plot
factor.wge(phi=c(1.9))                              # not invertible
factor.wge(phi=c(1.3,-.4))
```

```{r Prob 3.7d}
# Xt - 2.95Xt-1 + 3.87Xt-2.82Xt-3 +.92Xt-4 = at - .9at-1
plotts.true.wge(n=200,phi=c(2.95,-3.87,2.82,-.92),theta=c(.9)) # stationary
factor.wge(phi=c(2.95,-3.87,2.82,-.92))                        # invertible
factor.wge(phi=c(.9))
```

```{r Prob 3.7e}
# (1 - B - .49B^2 + .9B^3 - .369B^4)Xt = (1 + B + B^2 + .75B^3)at
plotts.true.wge(n=200,phi=c(2.95,-3.87,2.82,-.92),theta=c(.9)) # stationary
factor.wge(phi=c(1,.49,-.9,.369))                             
factor.wge(phi=c(-1,-1,-.75))                                  # invertible
```

```{r Prob 3.8a}
# Xt = at  - .5at-1 - 5at-2
plotts.true.wge(n=200,theta=c(.5,.5)) # stationary
factor.wge(phi=c(.5,5))              # not invertible             
```

```{r Prob 3.8b}
# Xt = at  - 2t-1 + 1.5at-2
plotts.true.wge(n=200,theta=c(2,-1.5)) # stationary
factor.wge(phi=c(2,-1.5))              # not invertible             
```
Show model is stationary and invertible then find the first 10 psi weight, remember the first is always 1
```{r Prob 3.11}
# (1-B-.49B^2 +.9B^3 - .369B^4)Xt = (1+B+B^2+.75B^3)at
plotts.true.wge(n=200,phi=c(1,.49,-.9+.369),theta=c(-1,-1,-.75)) # stationary
factor.wge(phi=c(1,.49,-.9,.369))                                # invertible
factor.wge(phi=c(-1,-1,-.75))
psi.weights.wge(phi=c(1,.49,-.9,.369,theta=c(-1,-1,-.75)),lag.max = 9)
```

```{r Prob 3.13i}
# (1 - 2.2B + 2.1B^2 - .8B^3)Xt = at
x=gen.arma.wge(n=100,phi=c(2.2,-2.1,.8),sn=1) 
x=factor.comp.wge(x,p=3,ncomp=2)            

```
Use least squares to fit a regression to find a and b (slope, intercept) 
Find and plot the residuals
Plot the autocorrelations
```{r Prob 5.1}
data(hadley)                         # load data
x=1:160                              # create an index
linreg=lm(hadley ~ x)                # fit a linear model, get slope and intercept
linreg                               # show slope and intercpet
linreg.res=resid(linreg)             # get residuals
plot(linreg.res,type="l")            # plot residuals
plotts.sample.wge(linreg.res)        # sample autocorrelations of residuals
plotts.sample.wge(hadley)            # sample autocorrelations of original data
```
Generate complex signal plus noise models
```{r Prob 5.2}
# Xt -.66Xt-1 + .02Xt-2 -.1Xt-3 -.24Xt-4 = at
gen.aruma.wge(n=150,phi=c(.66,-.02,.1,.24))                                        # AR(4)

# Xt = -0.526 + .0044t + Zt, where (1-.614B + .044B^2 - .077B^3 - .206B^4)Zt = 0
gen.sigplusnoise.wge(n=150,b0=-.526,b1=.0044,phi=c(.614,-.044,.077,.206),vara=.01) # linear + AR(4) + noise
```
ARIMA and ARUMA Models
```{r Prob 5.3}
# (1 - 1.2B + .8B^2)(1 - B)Xt = at
x=gen.arima.wge(n=200,phi=c(1.2,-.8),d=1)                        # stationary + 1 unit root 

# (1 - 1.2B + .08B^2)(1 - B)^2*Xt = at
x=gen.arima.wge(n=200,phi=c(1.2,-.8),d=2)                        # stationary + 2 unit roots

# (1 - 1.2B + .08B^2)(1 - B^4)Xt = at
x=gen.aruma.wge(n=200,phi=c(1.2,-.8),s=4,d=0,sn=1)               # stationary + 4th order seasonal 

# (1 - 1.2B + .08B^2)(1 - B)^2(1 - B^4)(1 + B + B^2)Xt = at
x=gen.aruma.wge(n=200,phi=c(1.2,-.8),s=4,d=2,lambda=c(-1,-1))    # stationary + 4th order seasonal + 2 unit roots
```

```{r Prob 5.5a Factor}
# (1 - 3B + 4.5B^2 - 5B^3 + 4B^4 - 2B^5 + .5B^6)Xt = (1 - 1.7B + .8B^2)at
factor.wge(phi=c(3,-4.5,5,-4,2,-.5))                                        # ARIMA(4,2,2)
factor.wge(phi=c(1.7,-.8))
```

```{r Prob 5.5b Factor}
# (1 - .5B + .3B^2 - .95B^3 + .3B^4 - .35B^5 + .2B^6)Xt = at
factor.wge(phi=c(.5,-.3,.95,-.3,.35,-.2))                                   # ARIMA(5,1,0)
#(1-B)(1+B+.8B^2)(1+.5B^2)(1-.5B)Xt = at
```

```{r Prob 5.5c Factor}
# (1 - 1.5B + 1.3B^2 + .35B^3 - B^4 + 1.35B^5 - .7B^6 + .4B^7)Xt = (1 + 0B + .9B^2)at
factor.wge(phi=c(1.5,-1.3,-.35,1,-1.35,.7,-.4))                             # ARUMA(6,1,2)
factor.wge(phi=c(0,-.9))
# (1+B)(1-1.5B+B^2)(1-B+B^2)(1+.5B^2)Xt = (1+.9B^2)at
```

```{r Prob 5.5d Factor}
# (1 - 3B + 4.5B^2 - 5B^3 + 4B^4 - 2B^5 + .5B^6)Xt = (1 - 1.7B + .8B^2)at
factor.wge(phi=c(3,-.4,5,5,-4,2,-.5))
factor.wge(phi=c(1.7,-.8))
```
Seasonal Plots
```{r Prob 5.7}
# (1 - B^6)Xt = at
factor.wge(phi=c(rep(0,5),1))
gen.aruma.wge(n=120,s=6)
x=plotts.true.wge(n=120,phi=c(rep(0,5),.99999))
plotts.sample.wge(x$data)
```

